Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2014-10-24 concepts.tex
added dependency definition

>>>>

Another fundamental concept, which adds further complexity to the
problem of multiobjective design space exploration, is the following

\begin{definition}[Parameter dependencies]
\defL{dependency}
A parameter $x$ is said to be dependent on another parameter $y$ if
the optimal value of $y$, i.e. the $x_{opt}$ which optimizes the
objectives is dependent upon the current value of $y$. 
\end{definition}
Note that when this happens, we cannot make any safe assumption on the
``meaning'' of $a$ in order to intuitively choose an appropriate
value.

### 2014-10-24 experiments.tex
added dependency definition

\section{A Case Study: VLIW Hardware/Software Design Space}
\secL{ee}
Since our main focus is the embedded system design, we apply PS to a
related case study.  In this section, we briefly describe the
parameterized platform used as testbed for the experiments, the
general evaluation flow along with the high-level estimation models
and applications used as benchmarks.

\subsection{Motivation}
In order to evaluate the behavior of the proposed approach
we use a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

\subsection{Parameter Space}
In general, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. 
The first is directly related to the VLIW core and includes the size
of the register files and the number of functional
units of each type. In particular, five different types of register
files are supported:
GPR (32-bit registers for integers), FPR (64-bit registers for
floating point values) PR (1-bit registers used to store the Boolean
values of predicated instructions), BTR (64-bit registers containing
information about possible future branches) and CR (32-bit control
registers containing information about the internal state of the
processor). The functional units involved are: \emph{Integer units},
\emph{floating point units}, \emph{memory units} (associated with
load/store operations) and \emph{branch units} (associated with branch
operations). With respect to the memory sub-system, the parameters
include \emph{size}, \emph{associativity} and
\emph{block size} for each of the three caches: First-level data cache
(L1D), first-level instruction cache (L1I) and second-level unified
cache (L2U).

Next, a set of the most impacting code compilation parameters has been
selected and included in the configuration space:
\begin{itemize}
\item \emph{tcc\_region}: Specifies the scope of action of the compiler and the type
of code transformation involved (basic block, super block and hyper
block) 
\item \emph {max\_unroll\_allowed}: The number of unroll iterations allowed
\item \emph{regroup\_only}: Avoids inlining 
\item \emph{do\_classic\_opti}: A set of classical optimization, not VLIW related,
such as common expression removal and loop unrolling
\item \emph{do\_prepass\_scalar\_scheduling}: Performs a schedule before
forming regions 
\item \emph{do\_postpass\_scalar\_scheduling}: Performs a schedule after the region formation 
\item \emph{do\_modulo\_scheduling}: Modular scheduling of loop code
\item \emph{memvr\_profiled}: Performs a memory-dependencies profiling 
\end{itemize}

>>>>

\section{A Case Study: VLIW Hardware/Software Design Space}
\secL{ee}
In order to evaluate the behavior of the proposed approach
we use a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

In this section, we will also briefly describe the parameterized platform used
as testbed for the experiments, the general evaluation flow along with
the high-level estimation models and applications used as benchmarks.

\subsection{Design Space: Hardware and Software Parameters}
Conceptually, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. The
first two categories are directly related to the physical
implementation of the system, so parameters of those
classes are usually referred ``hardware parameters''. On the other
side, all the parameters affecting the behaviour of the compiler
are referred as ``software parameters'', being directly involved in
the process of generating the application executables running on the
underlying hardware.

Processor hardware parameters are directly related to the VLIW core and includes the size
of the register files and the number of functional
units of each type. In particular, five different types of register
files are supported:
GPR (32-bit registers for integers), FPR (64-bit registers for
floating point values) PR (1-bit registers used to store the Boolean
values of predicated instructions), BTR (64-bit registers containing
information about possible future branches) and CR (32-bit control
registers containing information about the internal state of the
processor). The functional units involved are: \emph{Integer units},
\emph{floating point units}, \emph{memory units} (associated with
load/store operations) and \emph{branch units} (associated with branch
operations). With respect to the memory sub-system hardware, the parameters
include \emph{size}, \emph{associativity} and
\emph{block size} for each of the three caches: First-level data cache
(L1D), first-level instruction cache (L1I) and second-level unified
cache (L2U).

On the software side, a set of the most impacting code
compilation parameters has been selected and included in the
configuration space, in particular:
\begin{itemize}
\item \emph{tcc\_region}: Specifies the scope of action of the compiler and the type
of code transformation involved (basic block, super block and hyper
block) 
\item \emph {max\_unroll\_allowed}: The number of unroll iterations allowed
\item \emph{regroup\_only}: Avoids inlining 
\item \emph{do\_classic\_opti}: A set of classical optimization, not VLIW related,
such as common expression removal and loop unrolling
\item \emph{do\_prepass\_scalar\_scheduling}: Performs a schedule before
forming regions 
\item \emph{do\_postpass\_scalar\_scheduling}: Performs a schedule after the region formation 
\item \emph{do\_modulo\_scheduling}: Modular scheduling of loop code
\item \emph{memvr\_profiled}: Performs a memory-dependencies profiling 
\end{itemize}

### 2014-10-24 introduction.tex
added dependency definition

\section{Related work}
\secL{Related-work}
Many different design space exploration algorithms have been proposed
in literature.  The motivations of an exploration
algorithm are rather heuristic, have some form of arbitrariness and,
to a large extend, intuition lies behind them.

@@

A preliminary estimation of hardware/software dependencies in order to
reduce the design space is addressed in~\cite{Catania2008}, where
authors propose to deal with the software parameters separately,
statistically testing their average effect on objectives before the
actual exploration is performed. Also in~\cite{merging} authors
analyze the impact of including compilation parameters into the design
space space exploration. The results of both works confirm the
important role that hardware/software dependencies can play in a
multi-objective exploration and thus justify the choice of explicitly
including the two class of parameters in the design space without
making any distinction between them from the algorithm's perspective.

@@

Many studies, as \cite{coello_easmop} and others, focus on genetic
approaches to solve multiobjective optimization problems. Genetic
approaches have many advantages: they provide good accuracy, they are
customizable and very general and require no a-priori knowledge to the
designer. The strong point of genetic approaches can be summarized as
follows: they consist in an exploration that is sufficiently broad
(the mutation avoids being rigidly restricted to limited parts of
parameter space), and, at the same time, not too scattered as it is
guided by the performances of the already evaluated configurations.
Different variants of these algorithms have been proposed in
literature, including mixed approaches as in~\cite{Ascia2011382},
where authors specifically focuse on increasing the performance of a
evolutionaly exploration strategy, proposing a fuzzy based technique
to avoing the actual simulation of a certain number of configurations.
In particular, in this work we compare the proposed approach against a
well known Multi-Objective Genetic Algorithm~\cite{knowles_techrep06}
which is at best of our knowledge one of the most appropriate for the
particular scenario we adopted as testbed.  We claim that the approach
presented in this paper has most of the benefits of the genetic
approaches, although the rationale is completely different.

>>>>

\section{Related work}
\secL{Related-work}
Many different design space exploration algorithms have been proposed
in literature.  What makes the exploration of embedded systems so
complex is not only the huge size of the parameter spaces involved,
but also the tight interaction between parameters in relation to their
effect on the objective to be optimised, which take the form of
\emph{parameter dependencies} (see~\defR{dependency}). The 
motivations of an exploration algorithm and how it deals with such
complexity are often rather heuristic, have some form of arbitrariness and,
to a large extend, intuition lies behind them.

@@

A preliminary estimation of hardware/software dependencies in order to
reduce the design space is addressed in~\cite{Catania2008}, where
authors propose to deal with the software parameters separately,
statistically testing their average effect on objectives before the
actual exploration is performed. Also in~\cite{merging} authors
analyze the impact of including compilation parameters into the design
space space exploration. The results of both works confirm the
important role that hardware/software dependencies can play in a
multi-objective exploration and thus justify the choice of explicitly
including the two class of parameters in the design space without
making any distinction between them from the algorithm's perspective.
In fact, one of the purposes of the proposed PS algorithm is avoiding
the need of any a priori knowledge about those dependencies, letting
the algorithm itself discover regions where interaction between
parameters generates new pareto-optimal points (see the PS scoring
system described in the next sections).

@@

Many studies, as \cite{coello_easmop} and others, focus on genetic
approaches to solve multiobjective optimization problems. Genetic
approaches have many advantages: they provide good accuracy, they are
customizable and very general and require no a-priori knowledge to the
designer. The strong point of genetic approaches can be summarized as
follows: they consist in an exploration that is sufficiently broad
(the mutation avoids being rigidly restricted to limited parts of
parameter space), and, at the same time, not too scattered as it is
guided by the performances of the already evaluated configurations.
Different variants of these algorithms have been proposed in
literature, including mixed approaches as in~\cite{Ascia2011382},
where authors specifically focuse on increasing the performance of a
evolutionaly exploration strategy, proposing a fuzzy based technique
to avoid the actual simulation of a certain number of configurations.
In particular, in this work we compare the proposed approach against a
well known Multi-Objective Genetic Algorithm~\cite{knowles_techrep06}
which is at best of our knowledge one of the most appropriate for the
particular scenario we adopted as testbed.  We claim that the approach
presented in this paper has most of the benefits of the genetic
approaches, although the rationale is completely different.
